Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-04 linear_iso.tex
restructing, moving iso to appendix

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. We examine isothermal perturbations ($\beta =
0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

@@

\subsection{Governing equation in the VGSB}\label{approx_gov}
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}. %stoll kley see low
  % freq 
  We call this the low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks. % ,
% although further simplifications or maniputations will be performed
% depending the type of perturbations being considered.

@@

% We will consider sound-speed profiles with $q<0$ in the global 
% disk. Then $\Omega^2(z)$ decreases away from the midplane,
% and may reach small values for sufficiently large $|z|$, which
% invalidates the low-frequency approximation. In principle, then, the
% above approximations implicitly limits us to consider thin disks ($\epsilon
% \ll 1$) with vertical domain sizes not too large, so that the rotation
% is nearly-Keplerian.

@@

% Consider rotation profiles with radial dependence
% $\propto r^{-m}$. Then the same dimensionless governing equation 
% (Eq. \ref{vertiso_gov_nondim}) holds, provided we redfine $\khat$
% as $k_xc_s/\kappa$, $\hat{\sigma}$ as $\sigma/\kappa$, and reset
% $\epsilon$ to $\epsilon/sqrt{4-2m}$.

\subsection{Isothermal perturbations}\label{iso_discuss}
Isothermal perturbations correspond to $\beta\equiv0$ 
%relaxation timescale $\be
%or for any adiabatc index $\gamma$ when
%$\beta=0$. 
($\chi=1$) and Eq. \ref{ode_Q} becomes  
\begin{align}
  Q = W 
\end{align}
(i.e. $\delta P = c_s^2\delta \rho$), and so  
\begin{align}
  \ii\hat{\sigma}c_s\delta v_z = W^\prime
  % \quad   \text{(isothermal perturbations)},
\end{align}
from Eq. \ref{lin_vz}. For isothermal perturbations it is in fact 
simpler to work with $W$ by integrating Eq. \ref{vertiso_gov_nondim}. For later discussion
it is convenient to write the result as

\begin{align}
  0 = W^{\prime\prime} + \left[\ln\rho^\prime - \ii \epsilon q \hat{k}
    f(\zhat)\right]W^\prime + \hat{\sigma}^2\left(1+\hat{k}^2\right)W,\label{iso_ode1}
\end{align}
where $f(\zhat)$ is defined such that
\begin{align}\label{fz_shear}
  \frac{d\Omega^2}{d\hat{z}} = \epsilon^2q f(\hat{z})\Omega_k^2.
\end{align}
For a vertically isothermal disk we have 
\begin{align}
  f(\zhat) = \frac{\zhat}{\left(1+\epsilon^2\hat{z}^2\right)^{3/2}}, 
\end{align}
%For the vertically isothermal disk we have $|f(\hat{z})| =
%|\hat{z}|\left(1+\epsilon^2\hat{z}^2\right)^{-3/2}$, 
which maximizes at $|\zhat|=1/\sqrt{2}\epsilon$. The existence of a
maximum vertical shear limits the growth rate of unstable modes, as  
shown below.

Note that Eq. \ref{iso_ode1} can in fact be derived by substituting
$Q=W$ in Eq. \ref{ode_w} without approximating $D$ or $\kappa^2$ as
vertically constant.

% the issue of ignoring the
% vertical dependence of $D$ described in \S\ref{approx_gov} and Appendix
% \ref{adia_improve} is irrelevant, since
% Eq. \ref{iso_ode1} may be derived directly , i.e. no derivatives of $D$ needs to be taken.

%We remark that Eq. \ref{iso_ode1} is also applicable for  $\gamma=\Gamma\neq 1$ in the
%limit $t_c\to\infty$ (i.e. neutrally stratified adiabatic disks),
%since in that case $Q\simeq W$ as well. However, the following
%discussion is only  valid for $\gamma=\Gamma=1$ because $c_s$ is taken
%to be a constant.

\subsubsection{Necessity of vertical shear for
  instability}\label{integral_relation} 
We can establish a necessary condition for instability by multiplying
Eq. \ref{iso_ode1} by $\rho W^*$ and integrate vertically from
$\zhat=\zhat_1$ to $z=\zhat_2$. We neglect boundary 
terms when integrating by parts, by assuming $W$ or
$W^\prime$ vanishes at the boundaries, or that the boundaries are 
sufficiently far away so that the boundary terms are negligible because of the
decaying background density with increasing height. Then,
\begin{align}
  &\hat{\sigma}^2\left(1+\hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|W|^2d\zhat \notag\\
  &=\int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\zhat 
  +\ii \epsilon q \hat{k}\int_{\zhat_1}^{\zhat_2}\rho f(\zhat) W^*W^\prime d\zhat.\label{integral_relation1}
\end{align}
It follows that for instability ($\imag\hat{\sigma}>0$), it is necessary to
have $q\neq0$ or more generally $d\Omega^2/dz\neq 0$, i.e. vertical
shear. A similar conclusion can be reached in the equivalent linear  
problem in global cylindrical disk geometry.

\subsubsection{Maximum growth rate}   
Here we show that the growth rate is limited by the maximum vertical
shear in the domain. Writing $\hat{\sigma} = \hat{\omega} +
\ii\hat{\nu}$ with real $\hat{\omega}$ and $\hat{\nu}$, the real and
imaginary parts of 
Eq. \ref{integral_relation1} are
\begin{align}
  &\left(\hat{\omega}^2-\hat{\nu}^2\right)\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
  \int_{\zhat_1}^{\zhat_2}\rho|W^\prime|^2d\hat{z}
  \notag\\
  &=\real\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\zhat\right], \notag\\
  & 2\hat{\omega}\hat{\nu}\left(1+\hat{k}^2\right)
  \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z}\notag\\
  &=\imag\left[\ii
    \epsilon\hat{k}q \int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right].
\end{align}
Adding the square of these equations give
\begin{align}
  &\left[|\hat{\sigma}|^2\left(1+\hat{k}^2\right)
    \int_{\zhat_1}^{\zhat_2}\rho|W|^2d\hat{z} -
    \int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\right]^2\notag\\
  &+4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\notag\\
  &=\left|\ii
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
It is clear that
\begin{align}\label{sigma_finite_domain} 
  &4\hat{\nu}^2\left(1+\hat{k}^2\right) 
  \int_{\zhat_1}^{\zhat_2}\rho
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}\notag\\
  &\leq\left|
    \epsilon\hat{k}q\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|^2.
\end{align}
On the left hand side of this inequality, we apply the Cauchy-Schwarz
inequality to obtain
\begin{align}
  &\left( \int_{\zhat_1}^{\zhat_2}\rho
    |W|\left|W^\prime \right|d\hat{z}\right)^2\leq
  \int_{\zhat_1}^{\zhat_2}\rho 
  |W|^2d\hat{z}\int_{\zhat_1}^{\zhat_2}\rho\left|W^\prime \right|^2d\hat{z}.
\end{align}
On the right hand side of Eq. \ref{sigma_finite_domain} we have
\begin{align}
  \left|\int_{\zhat_1}^{\zhat_2}\rho
    f(\hat{z}) W^*W^\prime d\hat{z}\right|\leq \int_{\zhat_1}^{\zhat_2}\rho
  \left|f(\hat{z})W^*W^\prime \right|d\hat{z} \notag\\
  \leq
  \mathrm{max}\left(|f|\right)\int_{\zhat_1}^{\zhat_2}\rho
  |W|\left|W^\prime \right|d\hat{z},
\end{align}
where $\mathrm{max}(|f|)$ is the maximum value of $|f|$ in
$\zhat\in[\zhat_1,\zhat_2]$. Inserting these inequalities into
Eq. \ref{sigma_finite_domain} gives
\begin{align}\label{max_growth}
  |\hat{\nu}|\leq
  \frac{\epsilon |\hat{k} q|}{2\sqrt{1+\hat{k}^2}}\mathrm{max}(|f|). 
\end{align}
Recalling that $f$ represents vertical shear (Eq. \ref{fz_shear}), it
follows that the maximum possible growth rate of unstable modes,
satisfying the above boundary conditions, is limited by the maximum
vertical shear in the domain considered.

In practice, one might consider a vertical domain of a few scale 
heights in a thin disk with $|q|=O(1)$. In this case, $|\epsilon\zhat|\ll1$ and  
$f\simeq \zhat$, so that $\mathrm{max}|f| = O(1)$, implying a
maximum growth rate $O(\epsilon \Omega_k)$, consistent with numerical
results below.

% For vertical domains smaller 
% than this height, the growth rate is limited by $|f|$ at the vertical
% boundaries considered. For $|\epsilon\hat{z}|\ll1$ we have $f\simeq
% \hat{z}$, in which case we can expect growth rates to increase at most linearly
% with respect to the vertical domain size.   %no faster than linear

\subsubsection{Thin-disk limit}
We can make further progress for thin disks ($\epsilon\ll1$), 
by expanding the background density and vertical shear profile in powers
of $z/r$. To lowest order we obtain 
\begin{align}
  &\frac{\rho}{\rho_0} \simeq 
  \exp{\left(-\frac{\hat{z}^2}{2}\right)} \equiv w(\hat{z}),\label{thin_dens}\\
  &\frac{d\Omega^2}{d\hat{z}} \simeq \epsilon^2q\Omega_k^2\hat{z}, \label{thin_vshear}
\end{align} 
%With this approximation and those in \S\ref{integral_relation}, 
and the governing equation becomes 
\begin{align}\label{iso_ode3}
  W^{\prime\prime} - \left(1 + \ii q\epsilon
    \hat{k}\right)\hat{z}W^\prime  +
  \hat{\sigma}^2\left(1+\hat{k}^2\right)W = 
  0.
\end{align}
We remark that for large $|\hat{k}|$, Eq. \ref{iso_ode3} is the same as
that derived by \cite{nelson13}, although we have taken a different
route.

To complete the problem we must specify boundary conditions. The 
maximum vertical domain size should be limited by the thin
disk and low-frequency approximations (i.e. $z\lesssim r$).     
It is, however, common to take $\mathrm{max}|z|\to\infty$ and impose
that the kinetic  energy remain bounded at infinity. This 
is appropriate for the density field because both the true density profile 
and its thin-disk approximation decay rapidly away from the midplane.

However, the thin-disk representation of vertical shear diverges with 
$\hat{z}$, unlike the true vertical shear profile which eventually
decays.  Thus, taking $\zhat_{1,2}\to \pm\infty$ will permit unphysically large growth
rates (as demonstrated below). Nevertheless, the infinite disk is
useful to consider since it permits an analytical discussion.

\subsubsection{Stability in the absence of vertical shear}\label{iso_stable}
When $q=0$, Eq. \ref{iso_ode3} is
\begin{align}\label{hermite_ode}
  \left[w(\hat{z})W^\prime \right]^\prime + nW
  w(\hat{z}) =0, 
\end{align}
where we have defined a new eigenvalue
\begin{align}
  n \equiv \hat{\sigma}^2(1+\hat{k}^2). 
\end{align} 
Eq. \ref{hermite_ode} is Hermite's differential equation. If we impose
that the kinetic energy density remain bounded at infinity, then the
solutions are Hermite polynomials  
\begin{align}
  W \propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\hat{\sigma}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. \label{iso_stable_disp}
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2\sim \Omega_k^2$, our
analysis is only valid for large wavenumbers such that $\hat{k}^2\gg   
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. Note that
Eq. \ref{iso_stable_disp} is implied by Eq. 54 in
\cite{lubow93} after setting $\gamma=1$ and making the low-frequency,
Keplerian disk approximation in that equation.

We remark that for fixed $n$ and $\khat\gg 1$ the dispersion relation 
Eq. \ref{iso_stable_disp} imply $\hat{\omega}\propto 1/\khat$. This inverse relation 
has also been qualitatively observed in numerical simulations
\citep{stoll14}.

\subsubsection{Polynomial solutions with vertical shear}\label{iso_poly}
We can seek power-series solution to Eq. \ref{iso_ode3},
\begin{align}
  W(\zhat) = \sum_{l=0}^\infty a_l\zhat^l. 
\end{align}
Then the coefficients must satisfy the recurrence relation
\begin{align}
  (l+2)(l+1)a_{l+2} +
  \left[n - l\left(1+\ii \epsilon q  \hat{k}\right)\right] a_l = 0, 
\end{align}
where $n$ is not necessarily an integer. Indeed, if we demand
a polynomial of degree $L$ as the solution, then the eigenfrequency
must satisfy
\begin{align}\label{sig2_iso}
\hat{\sigma}^2 = L\left(\frac{1+\ii q \epsilon
    \hat{k}}{1+\hat{k}^2}\right).
\end{align}
The eigenfrequency $\hat{\sigma}$ is therefore complex for
$L\geq1$. We are interested in 
unstable modes for which $\hat{\nu}=\imag{\hat{\sigma}}>0$ is the
growth rate and is given by 
\begin{align}\label{simple_growth}
  \hat{\nu} =\sqrt{
   \frac{L}{2\left(1+\hat{k}^2\right)}\left(\sqrt{1+q^2\epsilon^2\hat{k}^2} - 
    1\right)}. 
\end{align}
For fixed $q$ and $\epsilon$, the growth rate vanishes for both
$\hat{k}^2\to0$ and $\hat{k}^2\to\infty$. 
The maximum growth rate
occurs at the optimum wavenumber $\hat{k}_\mathrm{opt}$
\begin{align}
  |\hat{k}_\mathrm{opt}| = \sqrt{\frac{2+|\epsilon q|}{|\epsilon q|}},
\end{align}
and
\begin{align}
  \mathrm{max}\left(\hat{\nu}\right) =\frac{\sqrt{L}|\epsilon
    q|}{2\sqrt{1+|\epsilon q|}}. \label{iso_max_growth}
\end{align}
Since $L\geq1$, there is no limit to the maximum growth rate. This is
an artifact of the thin-disk approximation because the vertical 
shear $d\Omega^2/dz\propto z$ increases indefinitely with
height. However, we find Eq. \ref{iso_max_growth} compares well with
numerical solutions of the governing equation for small $L$.

>>>>

\section{Analytical discussion of vertically isothermal
  disks in the framework of the vertically global shearing
  box}\label{analytical}  
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the radially-local
approximation. % We examine isothermal perturbations ($\beta =
% 0$) before considering the effect of finite thermal relaxation. 
In order to facilitate a fully analytic discussion, we first introduce a
number of approximations below that allows a simple governing equation
to be derived and compared with previous works.

@@

\subsection{Low-frequency and nearly-Keplerian
  approximation}\label{approx_gov} 
We now derive a governing equation with further approximations:
\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}. %stoll kley see low
  % freq 
  We call this the low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify 
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with
  $\Omega_k^2$. We call this the nearly-Keplerian approximation.

@@

It is convenient to introduce the dimensionless variables
\begin{align}
  \hat{z} = z/H,\quad \hat{k}=k_xH, \quad \hat{\sigma} = \sigma/\Omega_k,
\end{align}
then Eq. \ref{vertiso_gov} becomes
\begin{align}
  0=& \delta v_z ^{\prime\prime} + \left(1 + \ii\epsilon
    q\hat{k}\right)\ln\rho^{\prime}\delta v_z^\prime \notag\\
  &+
  \left[\hat{\sigma}^2\left(\hat{k}^2+\chi\right) +
    \left(  \chi + \ii \epsilon q\hat{k}\right)\ln\rho^{\prime\prime}\right.
  \notag\\
  &\left.- \ln\rho^{\prime
      2}\left(\khat^2 -
      \ii\epsilon
      q\hat{k}\right)\left(1 - \chi\right)\right]\delta v_z,\label{vertiso_gov_nondim}
\end{align}
where $^\prime$ denotes $d/d\zhat$. Eq. \ref{vertiso_gov_nondim} will
be the basis for our discussion of vertically isothermal disks.

@@

@@

### 2015-03-04 setup.tex
restructing, moving iso to appendix

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot}, but is also
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}

@@

>>>>

Similarly, the equilibrium rotation is 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\Omega_k^2(r)}=1 +
  \left(p+\frac{s}{\Gamma}\right)\epsilon^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
where $s\equiv q+p(1-\Gamma)$. We also define % the radial shear rate  
% \begin{align}
%   S(r,z) \equiv r\frac{\p\Omega}{\p r},  
% \end{align}
% and
the square of the epicyclic frequency $\kappa$ 
\begin{align}
  \kappa^2(r,z) \equiv \frac{1}{r^3}\frac{\p }{\p
    r}\left(r^4\Omega^2\right). %2\Omega(2\Omega + S). 
\end{align}

@@

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
Evaluating this expression explicitly,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\Omega_k^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that there is a maximum vertical shear rate at $z=r/\sqrt{2}$,
which is expected to limit the growth rate of unstable modes.
